---
title: 'Computer practical 1-cont: Topics in Statistics III/IV, Term 1'
author: "Georgios Karagiannis"
output:
  html_notebook: default
  html_document: default
  word_document: default
  pdf_document: default
---

<!-- -------------------------------------------------------------------------------- -->

<!-- Copyright 2020 Georgios Karagiannis -->

<!-- This file is part of Topics_in_Statistics_Michaelmas_2020 -->
<!-- (Topics in Statistics III/IV (MATH3361/4071) Michaelmas term 2020) -->
<!-- which is the material of the course (Topics in Statistics III/IV (MATH3361/4071) -->
<!-- taught by Georgios P. Katagiannis in the Department of Mathematical Sciences   -->
<!-- in the University of Durham  in Michaelmas term in 2020 -->

<!-- Topics_in_Statistics_Michaelmas_2020 is free software: you can redistribute it and/or modify -->
<!-- it under the terms of the GNU General Public License as published by -->
<!-- the Free Software Foundation version 3 of the License. -->

<!-- Topics_in_Statistics_Michaelmas_2020 is distributed in the hope that it will be useful, -->
<!-- but WITHOUT ANY WARRANTY; without even the implied warranty of -->
<!-- MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the -->
<!-- GNU General Public License for more details. -->

<!-- You should have received a copy of the GNU General Public License -->
<!-- along with Topics_in_Statistics_Michaelmas_2020 If not, see <http://www.gnu.org/licenses/>. -->

<!-- -------------------------------------------------------------------------------- -->


> Aim
>
> * To check some descriptive statistics:   
>   * Fourfold Plots  and Mosaic Plots


---------

# Contigency table: data manipulation

Below we load a table where refers to a 1992 survey by the Wright State University School of Medicine and the United Health Services in Dayton, Ohio. The survey asked 2276 students in their final year of high school in a nonurban area near Dayton, Ohio whether they had ever used alcohol, cigarettes, or marijuana. Denote the variables in this $2 \times 2 \times 2$ table by A for alcohol use, C for cigarette use, and M for marijuana use.

> Load the observed counts in a data frame ` obs.frame  ` and print the result. Use commands :
> 
> * `data.frame()` : as in SC2
> * `factor()` :  to encode a vector as a factor (aka category)
> * `expand.grid()` : to produce all combinations of the supplied vectors or factors.


```{r}

# I will do this for you

## load the data

obs.frame<-data.frame(count=c(911,538,44,456,3,43,2,279),
                      expand.grid( 
  marijuana=factor(c("Yes","No"),levels=c("No","Yes")),
  cigarette=factor(c("Yes","No"),levels=c("No","Yes")),  
  alcohol=factor(c("Yes","No"),levels=c("No","Yes")))
  ) 

## orint the obs.frame

obs.frame

```

> Create $3$ dimentional contingency table from `obs.frame`. Use command:
> 
> * `xtabs()`, to create a contingency table from cross-classifying factors in a dara.frame

```{r}
# this is me again
obs.xtabs <- xtabs(count ~ marijuana+cigarette+alcohol, data=obs.frame)
## print 
obs.xtabs
```

> Compute the marginal contigency table of marijuana and cigarette. 
> 
> *  Use command `margin.table( , margin = )` and `obs.xtabs`. 
> 
> *  Save it in `obs.mc.xtabs`.

```{r}
obs.mc.xtabs <- margin.table(obs.xtabs, margin=c(1,2))
obs.mc.xtabs
```


---

# Odds ratio calculations

> Code an R function, named 'odds.ratio' with:  
>
> * Inputs:  
>   * x : a 2 by 2 matrix whose elements are the observed counts of a 2 by 2 contigency table  
>   * conf.level : with default input value 0.95 representing the confidence level  
>   * theta0 : with default value 1 representing a null hypothesis value of the odds ratio test  
>
> * Outputs:  
>   * estimator : representing mle of odds satio  
>   * log.estimator : representing the mle of log odds ratio
>   * asympt.SE : representing the standard error / standard deviation of the mle of odds ratio  
>   * conf.interval: representing  confidence interval of mle of odds ratio at sig level conf.level (from the inputs)
>   * conf.level =representing  confidence level 
>   * Ztest : representing the test statistic for the odds ratio test (2 tails) 
>   * p.value : representing the p value of the odds ratio test (2 tails) 
>   * log.conf.interval : representing in log scale the confidence interval of mle of odds ratio at sig level conf.level (from the inputs)


```{r}
odds.ratio <- function(x,conf.level=0.95,theta0=1)
{
  if (any(x==0)) x <- x+0.5
  theta <- x[1,1] *
    x[2,2]/(x[1,2]  *
              x[2,1])
  SE <- sqrt(sum(1/x))
  Za2 <- qnorm(0.5 *
                 (1+conf.level))
  Low <- exp(log(theta)-Za2 * SE)
  Up <- exp(log(theta)+Za2  *
              SE)
  CI <- c(Low,Up)
  Z=(log(theta)-log(theta0))/SE
  pv=2   *
    pnorm(-abs(Z))
  
  logCI <- log(CI)
  
  list (estimator=theta, 
        log.estimator=log(theta), 
        asympt.SE=SE, 
        conf.interval=CI,
        conf.level=conf.level, 
        Ztest=Z, 
        p.value=pv,
        log.conf.interval=logCI
        )
}
```

> For the  marginal contigency table of marijuana and cigarette,  
>
> * compute the mle of the marginal odds ratio of marijuana and cigarette  
> * computet the 95% Confidence Interval of the marginal odds ratio of marijuana and cigarette  
> * perform a statiastical hypothesis test that marijuana and cigarette are independent at sig level 0.05 

```{r}
obs.mc.xtabs <- margin.table(obs.xtabs, margin=c(1,2))

odds.ratio.marijuana.cigarette <- odds.ratio(obs.mc.xtabs, conf.level=0.95, theta0=1 )
odds.ratio.marijuana.cigarette

```
The MLE of the marginal odds ratio of marijuana and cigarette  is `r  odds.ratio.marijuana.cigarette$estimator` .  

The 95% confidence interval of the marginal odds ratio of marijuana and cigarette is [`r odds.ratio.marijuana.cigarette$conf.interval`] .

The hypothesis test with $H_0:\theta=1$ versus $H_1:\theta\ne1$ at sig. level 0.05 has a p-value `r odds.ratio.marijuana.cigarette$p.value` Hence I reject $H_0:\theta=1$ against $H_1:\theta\ne1$ at sig. level 0.05. 


---

## Fourfold Plots

You can draw Fourfold Plots


### Tables 2 x 2

It is a graphical expression visualizing the odds ratio  

$$\theta = \frac{n_{11}n_{22}}{n_{12}n_{21}}$$

in 2 x 2 contingency tables. 

It shows the departure from independence as  measured by the sample odds ratio, 

Each cell $n_{ij}$ is represented as a quarter-circle with radius proportional to $\sqrt{n_{ij}}$ and area proportional to $n_{ij}$.  

* If there is no association $\theta=1$ between classification variables, the quarter-circles should form a circle.

* If there is positive association $\theta>1$ between classification variables, the  diagonal areas are greater  than the off-diagonal areas

* If there is negative association $\theta<1$ between classification variables, the  diagonal areas are smaller  than the off-diagonal areas

R provides a function to draw this kind of plots  by using the function `fourfoldplot' from the package `vcd' 

> * Install `vcd' package and load it 

```{r}
# install.packages('vcd') # IF NOT ALREADY INSTALLED ON YOUR PC, THEN UNCOMMENT AND RUN THIS COMMAND
library(vcd)
```

> Check in help the function fourfoldplot by usign the command `?fourfoldplot'
> 
> * Draw a Fourfold Plot for the marginal contigency table of marijuana and cigarette.  
> * Discuss what you can see  

```{r}
#
#
#
```

Note that:  
* The area of each shaded quadrant shows the observed counts.  
* Circular arcs show the limits of confidence interval for the odds ratio.  


### Tables 2 x 2 x K

Fourfold Plots can be also used for 2 x 2 x K contigancy tables 

> * Draw a Fourfold Plot for the contigency table of marijuana cigarette, and alcohol by controlling on the alcohol levels.  
> * inspect the plots

```{r}
#
#
#
```

## Mosaic plot 

Mosaic plot display graphically the cells of a contingency table as rectangular areas of size proportional to the corresponding observed frequencies.

When the classification variables are independent the areas tend to be perfectly aligned in rows and columns.  

The greater the deviation is, the worse the aforesaid alignment is.

Furthermore, specific locations of the table that deviate from independence the most can be identified and thus the pattern of underlying association can be explained. 

The strength of individual cells contribution to divergence from independence as well as the direction of the divergence are reflected in the magnitude and sign of the corresponding independence model’s residuals that can be incorporated in a mosaic plot.


R provides a function to draw this kind of plots  by using the function `mosaic' from the package `vcd' 

> * Install `vcd' package and load it 

```{r}
# install.packages('vcd') # IF NOT ALREADY INSTALLED ON YOUR PC, THEN UNCOMMENT AND RUN THIS COMMAND
library(vcd)
```

> * Check the command 'mosaic' in help by typing '?mosaic' 

> For the I x J  case:  
>  
> * Draw a Mosaic Plot for the marginal contigency table of marijuana cigarette.   
> * in particular use use mosaic(x,residuals_type="deviance",gp=shading_hcl)  where x is the contigency table of interest  
> * Interpretet the plots 


```{r}
#
#
#
```


> For the I x J x K  case:  
>  
> * Draw a Mosaic Plot for the contigency table of marijuana cigarette and alcohol.   
> * Interpretet the plots   

```{r}
#
#
#
```






---

# Save me  

Generate the document as a Notebook, PDF, Word, or HTML by choosing the relevant option (from the pop-up menu next to the Preview button). Then save your Markdown code by choosing the relevant option (from the task bar menu).

Save the *.Rmd script, so that you can edit it later.
