---
title: 'Computer practical 2: Topics in Statistics III/IV, Term 1'
author: "Georgios Karagiannis"
output:
  html_notebook: default
  html_document: default
  pdf_document: default
  word_document: default
---

> Aim
> 
> * To perform Fisher exact test  
> * To Mantel-Haenszel test  
>
> * To analyse contigency tables with Log-linear models
>    + Fit the model
>    + Perform inference
>    + Specify the Identifiability constraint (Contrasts)
>    + Perform model comparison (Nested, or Non-nested models)

---------

***Load the libraries***

```{r}
library(MASS) # for contigency tables and log-linear models
library(vcd) # to visualise categorical data
```


---------

# Fisher's Exact test

***Description***

To perform FIsher's exact test we can use the command `fisher.test {stats}`, as,

* `fisher.test(x)` 

where `x` is either a two-dimensional contingency table in matrix form, or a factor object. 

For the output of the command, please chck help `?fisher.test` 

***Dataset***

[Cancer dataset] The table below shows the results  of a retrospective  study comparing radiation therapy with surgery in treating cancer of the larynx. The response indicates  whether  the  cancer  was  controlled  for  at  least  two  years following treatment.

```{r}
# load the data
## I will do this for you

Cancer.frame<-data.frame(count=c(21,2,15,3),
                      expand.grid( 
                      Cancer=factor(c("Controled","Not-controled"),
                                    levels=c("Controled","Not-controled")),
                      Therapy=factor(c("Surgery","Radiation"),
                                     levels=c("Surgery","Radiation"))
                      )
  ) 

## print the obs.frame
Cancer.frame
```

***Question***

Create a matrix `Cancer.xtabs` from `Cancer.frame` by using `xtabs()`.  Print the result on the screen.

***Answer***

```{r}




```

***Question***

Perform Fisher's Exact test   

* Recall the type of the command's argument, and tranform it suitably by using `xtabs()`

***Answer***

```{r}



```

***Question***

Report and interpret the P-value for Fisher’s exact test whose alternative hypothesis is that the odds ratio is larger than $1$.  

$$
H_0: \theta=1 \  \text{ versus } \  H_1: \theta>1
$$ 

***Answer***

Write your discussion here ...
.  
.  
.  
.  
.  
.  
.  
. 



***Question***

Explain  how the P-values are calculated for this particular test. (you can check your notes: Handout: Contigency tables)

***Answer***

Write your discussion here ...
.  
.  
.  
.  
.  
.  
.  
. 



---------

# Chi-squared Test


***Description***

To perform Chi-squared Test on a contigency tables, we can use the command `chisq.test()` from the package `stats`, as,

* `chisq.test(x,...)` 

where `x` a numeric vector or matrix holding the data

For the output of the command, see in the help page `?chisq.test` 


***Dataset***

Use the [Cancer dataset]


***Question***

* Perform the Chi-squared Test in order to test if Cancer control and Therapy are independent. 

* What is the conclusion?

***Answer***

```{r}




```

Write your discussion here ...
.  
.  
.  


---------

# Mantel–Haenszel test

***Description***

We wish to perform Mantel-Haenszel chi-squared test of the null that two nominal variables are conditionally independent in each stratum, assuming that there is no three-way interaction.  

We can use the command  `mantelhaen.test(x,...)`, as,  

*  mantelhaen.test(x,alternative="two.sided")

where `x` is a 3-dimensional contingency table in array form where each dimension is at least 2 and the last dimension corresponds to the strata.

For more options check help `?mantelhaen.test`


***Dataset***

[Marijuana data-set] Consider the following table where refers to a 1992 survey by the Wright State University School of Medicine and the United Health Services in Dayton, Ohio. The survey asked 2276 students in their final year of high school in a nonurban area near Dayton, Ohio whether they had ever used alcohol, cigarettes, or marijuana. Denote the variables in this $2 \times 2 \times 2$ table by A for alcohol use, C for cigarette use, and M for marijuana use.


```{r}
# I will do this for you

## load the data

marijuana.frame<-data.frame(count=c(911,538,44,456,3,43,2,279),
                      expand.grid( 
  marijuana=factor(c("Yes","No"),levels=c("No","Yes")),
  cigarette=factor(c("Yes","No"),levels=c("No","Yes")),  
  alcohol=factor(c("Yes","No"),levels=c("No","Yes")))
  ) 

## print the obs.frame

marijuana.frame
```


***Question***

Test the hypothesis that the use of Marijuana and the use of cigarete are independent accross the levels of the use of alcohol, by using Mantel–Haenszel test.  

Report the result of your inference.

If you wish, you can play with the alternative hypothesis (see the help page)


***Answer***

```{r}



```

```{r}



```


Write your discussion here ...
.  
.  
.  



---------


# Log-linear models

## Data

Consider the [Marijuana data-set] used previously

```{r}
marijuana.frame
```


## Fit the Log-linear model

Fitting  a  loglinear  model  can  be  done  using  Iterative  Proportional  Fitting  (`loglm` of the package `MASS`)  or  Newton  Raphson (`glm` with poisson family, of the package `stats`).   The former uses  `loglm` from `MASS`.   `loglm`     accepts as  input    table output, crosstabs output (`xtabs` in R), or a formula using variables from a data frame.

### Specify  a model
 
Below is an example for fiting the Log-linear model [ACM]. Notice that [ACM] it can be writen in different ways. I set the arguments  `fit` and `param` to `T`  so that I can get fitted values and parameter estimates, as we see later. 

```{r}

fitAC.AM.CM_ <- loglm(count~1+alcohol+cigarette+marijuana
                      +alcohol:cigarette
                      +cigarette:marijuana
                      +alcohol:marijuana
                      ,data=marijuana.frame,
                      param=T,fit=T) # ACM 

fitAC.AM.CM <- loglm(count ~ alcohol*cigarette
                +alcohol*marijuana
                +cigarette*marijuana,
                data=marijuana.frame,
                param=T,fit=T) # ACM 

fitAC.AM.CM_

fitAC.AM.CM

```

***Question***

Notice the details in the two equivalent executions.  

Regarding the `formula` above, what is the relation between terms (1, alcohol, cigarette, alcohol:cigarette) and (alcohol*cigarette) ?

***Answer***


Write your discussion here ...
.  
.  
.  


### Update a model

Based on a given model, we can build, new ones. E.g., the [ACM] and the [AC,M],

```{r}
fitACM<-update(fitAC.AM.CM,
                .~. + alcohol:cigarette:marijuana) # ACM 

fitAC.M<-update(fitAC.AM.CM, 
                .~. - alcohol:marijuana - cigarette:marijuana) # AC, M

```

***Question***

Fit models [AM, CM], and [ A, C, M ] and print the output.


***Answer***

```{r}





```


***Question***

Get  the estimates  of  the  model  parameters  (the  lambdas) by applying `...$param` to the output object of the `loglm` or `update`.  

Get the fitted values by applying `...$fitted` the command `fitted()` to the output object of the `loglm` or `update`.

***Answer***

```{r}



```



## Inference

Likelihood ratio chi-squared test statistics are output using the `summary()` function for `loglm`. The functin provides:

* The formula of the model as `Formula:`
* The design matrix X in the vectorized form of the log-linear model $\log(\mu)=X\beta$ as `attr(,"factors")`
* The Pearson and Likelihood ratio Goodness-of-fit test



***Question***

Apply `summary()` function on the output object of the function `loglm` for the homogeneous association model  object. 

Does the data-set support the homogeneous association model at sig. level $5\%$ ?

***Answer***

```{r}



```


Write your discussion here ...
.  
.  
.  


# Stepwise model selection

Two popular procedures can be used in order to perform Variable selection in linear models. By saying variable here, we mean the factors of the log-linear model, or (equivalently) the classifier variables of the associated contigency table. They are the follwoing:

* Forward selection.  
  + Forward  selection  adds  terms  sequentially  until  further  additions  do  not
improve the fit. E.g., [A,C,M]->[AC,M]->[AC,AM]->... At each stage it selects the term giving the greatest improvement  in  fit. The selection criterion can be the minimum P-value  for  testing  the  term  in  the  model. Then the procedure may stop when  adding more terms may result in insignifical p-values.

* Backward elimination. Backward   elimination   begins   with   a   complex   model   and   sequentially
removes  terms until we reach a simple model such as reducing it further can will not fit the data well.  At  each  stage,  it  selects  the  term  for  which  its  removal  has the  least  damaging  effect  on  the  model   e.g.,  largest P-value.  It stops at the model such that removal of any of its term lead to rejection of the Goodness-of-Fit test.

There is not a commonly accepted answer to the question which procedure is better.  Both are okay and in use... What is your personal opinion? Discuss it with your fellow students; alternatively ask the instructor. 




## Model comparison among a set of nested models

***Comment***

Comparison of nested models can be done using the anova  method. It gives  the  likelihood  ratio  tests  comparing  hierarchical   loglinear  models  given  in  the  list  of  arguments. For example,  

```{r}
anova(fitAC.M, fitAC.AM.CM)
```

The column `Deviance` is the Devience...(i.e., the Likelihood ratio statistic of the model vs the saturated)

Regarding the likelihood ratio test of  

* [AC, M] vs. [AC,AM,CM] , the statistic is equal to $843.4526577$, the degrees of freedom are equal to $2$ and the p-value equal to $0.00000$

*  [AC,AM,CM] vs. [ACM], the statistic is equal to $0.3739859$, the degrees of freedom are equal to $1$ and the p-value equal to $0.54084$


***Question***

Compare models [ACM], [AC,AM,CM], [AM, CM], [AC, M], [A,C,M]. What is your conclusion?


***Answer***

```{r}


```


Write your discussion here ...
.  
.  
.  

## Model comparison among a set of non-nested models

Select a subset of predictor (classifier) variables from a larger set (e.g., stepwise selection) is a controversial topic. 

You can perform stepwise selection (forward, backward) using the `stepAIC( )` function from the `MASS` package. `stepAIC( )` performs stepwise model selection by exact AIC.

***Comment ***

Model selection based on either Forward selection, or Backword elimination procedures can be perforemed by using AIC as a criterion. 

The command to perofrm this procedure automatically is `stepAIC()`.

* Forward selection, starting from fitA.C.M: `stepAIC(fitA.C.M, direction = "forward")$anova`  

* Backward elimination, starting from fitACM: `stepAIC(fitACM, direction ="backward")$anova`

The `...$anova` is because we are interested in the `...$anova` return value only.

***Question ***

Apply Forward selection, and Backward elimination by using AIC.  

* What is your conclusion about the preferable model by using each method? 

* Do the two procedures produce the same result?

***Answer ***

```{r}



```


```{r}



```



Write your discussion here ...
.  
.  
.  




# Practice at home

## Data
 
Use the Marijuana data, and consider additional classifier variables Gender (G), and race (R). 

The dataset is given below.

```{r}

marijuana_new.frame<-data.frame(count=c(405,13,1,1,268,218,17,117,453,28,1,1,228,201,17,
                                        133,23,2,0, 0,23, 19,1,12,30,1,1,0,19,18,8,17),
                                expand.grid(cigarette=c("Yes","No"), 
                                            alcohol=c("Yes","No"),marijuana=c("Yes","No"),
                                            sex=c("female","male"),  
                                      race=c("white","other"))
                                ) 
marijuana_new.frame
```

***Question***

* Perform a model selection (by using the procedure of your preference) in order to find the best model that prepresents the dependences of the variables.  

* For the selected model, compute the fitted values, and the estimations of the parameters.

* Discuss the conclusions of your inference. 


*** Answer ***


```{r}




```

```{r}



```



Write your discussion here ...
.  
.  
.  



# Save me  

Generate the document as a Notebook, PDF, Word, or HTML by choosing the relevant option (from the pop-up menu next to the Preview button). Then save your Markdown code by choosing the relevant option (from the task bar menu).

Save the *.Rmd script, so that you can edit it later.
